From the function-sheaf dictionary to quasicharacters of $p$-adic tori

### 2015-07-21 dictionary.tex
Reading Section 4.

\begin{example}\ref{eg:H2}
Consider the simplest non-trivial case, where $\bG = \{1, i, j, k\} \cong \ZZ/2\ZZ \times \ZZ/2\ZZ.$  Using eqref{eq:H2comp},
we have $\Hh^2(\bG, \EEx) \cong \ZZ/2\ZZ$, on which $\Weil{}$ must act trivially, regardless of its action on $\bG$ itself.
The non-trivial element corresponds to the extension
\begin{equation} \label{eq:Qdef}
1 \to \EEx \to Q \to \bG \to 1,
\end{equation}
where $Q = \{c + c_ii + c_jj + c_kk \tq c,c_i,c_j,c_k \in \EE \mbox{ with exactly one nonzero}\}$ is a subgroup of the quaternion algebra over $\EE$.
Let $a$ be a $2$-cocycle corresponding to this extension, with values in $\{\pm 1\}$. When $\Frob{G}$ acts trivially on $\bG$,
any homomorphism $b : \bG \to \EEx$ will satisfy \eqref{eq:nohom}, and the corresponding $\alpha \oplus \beta$ are non-cohomologous in $\Hh^2(E_G^\bullet)$.
When $\Frob{G}$ exchanges $i$ and $j$, then we may take $b(1) = 1, b(i) = -1$ and $b(j) = b(k) = \pm 1$, up to coboundaries.
Finally, when $\Frob{G}$ cycles $i$, $j$ and $k$, any homomorphism $b : \bG \to \EEx$ will satisfy \eqref{eq:nohom},
but now the corresponding $\alpha \oplus \beta$ are all cohomologous in $\Hh^2(E_G^\bullet)$.  In each case, we may produce
an explicit character sheaf from the listed $a$ and $b$.

@@

Let $R$ be the ring of integers of $K$.
Write $\m$ for the maximal ideal of $R$ and set $R_n = R/\m^{n+1}$ for every non-negative integer $n$.
Suppose $X_K \in \mathcal{N}_K$ has N\'eron model $X$ over $R$.
Note that $X(K) = X(R)$.
A \emph{quasicharacter of $X(K)$} is a group homomorphism $X(K) \to \EEx$ that factors through $X(R) \to X(R_n)$ for some non-negative integer $n$.
We note that this definition is compatible with \cite{cassels-frohlich:AlgebraicNumberTheory}*{Ch XV, \S 2.3}.
The group of quasicharacters of $X(K)$ will be denoted by 
$\Hom_\text{}(X(K),\EEx)$
 and the subgroup of those that factor through $X(R_n)$ will be denoted by $\Hom_n(X(K),\EEx)$.
In this section we will see how to geometrize and categorify quasicharacters of $X(K)$ using character sheaves.

@@

For any $n\leq m$,  the surjective ring homomorphism $R_{m} \to R_n$ determines a
natural transformation 
\[
\varrho^R_{n\leq m} : \Gr^R_{m} \to \Gr^R_n
\]
between additive functors.
Crucially, $\varrho^R_{n\leq m}(X): \Gr^R_{m}(X)\to \Gr^R_n(X)$ is an affine morphism of $\Fq$-schemes, for every $R$-scheme $X$ and every $n\leq m$ \cite{bertrapelle-gonzales:Greenberg}*{Prop 4.3}.
This observation is key to the proof that, for any scheme $X$ over $R$, the projective limit 
\[
\Gr_R(X) \ceq \varprojlim_{n\in \NN} \Gr^R_n(X),
\]
taken with respect to the surjective morphisms $\varrho^R_{n\leq m}(X) : \Gr^R_{m}(X) \to \Gr^R_n(X)$,
exists in the category of group schemes over $\Fq$;
see \cite{EGAIV3}*{\S 8.2}.
This leads to the definition of what we shall call the {\it Greenberg transform}:
\[
\Gr_R : \Sch{R} \to \Sch{\Fq}.
\]
By construction, the $\Fq$-scheme $\Gr_R(X)$ comes equipped with morphisms 
\[
\varrho^R_n(X) : \Gr^R(X) \to \Gr^R_n(X),\qquad \forall n\in \NN.
\]

@@

%Let $X \in \mathcal{N}$ be as above.
Set $S = \Spec{R}$ and $S_n = \Spec{R_n}$;
note that $S_0 = \Spec{\Fq}$ is the special fibre of $S$.
Let $X$ be a smooth commutative group scheme over $S$.
%We now define a category which will geometrize quasicharacters of $X(K)$.
For every non-negative integer $n$, the Greenberg transform $\Gr^R_n(X)$ is a {\it smooth} commutative group scheme over $S_0$
%with component group scheme $\pi_0(X)_{\Fq} \ceq \pi_0(X) \times_S S_0$ and $\Fq$-rational points $X(R_n).$
%
The Greenberg transform $\Gr_R(X)$ of $X$ is a commutative group scheme over $\Fq$
%with component group scheme $\pi_0(X)_{\Fq}$
with $\Fq$-rational points $X(R)$.
%
The morphism of $\Fq$-schemes $\varrho^R_n(X) : \Gr_R(X) \to \Gr^R_n(X)$ induces a functor
\[
\varrho^R_n(X)^* : \CS(\Gr^R_n(X)) \to \CS(\Gr_R(X)),
\]
as in Lemma~\ref{lem:pullback}.
%This functor is faithful, so we may view $\CS(\Gr^R_n(X))$ as a subcategory of $\CS(\Gr_R(X))$.

@@

\begin{remark}
We offer the following alternate construction of $\QCS(X)$.
As above, let $X$ be a smooth group scheme over $R$.
Although $\Gr_R(X)$ is not locally of finite type and therefore not smooth,
let us consider the rigid monoidal category $\CS(\Gr_R(X))$ as defined in Section~\ref{ssec:category} without insisting that the commutative group $\Fq$-scheme $G$ is smooth.
%Let $X$ be a smooth commutative group scheme over $R$, with $R$ as above.
A quasicharacter sheaf for $X$ is an object of
the following rigid monoidal subcategory of $\CS(\Gr_R(X))$, denoted by $\QCS(X)$:
\begin{enumerate}
\item
objects in $\QCS(X)$ are the $\ell$-adic sheaves $\varrho^R_n(X)^*\cs{L}$, for $n\in \NN$ and $\cs{L} \in \CS(\Gr^R_n(X))$; 
\item
morphisms $\varrho^R_n(X)^*\cs{L} \to \varrho^R_m(X)^*\cs{L}'$ in $\QCS(X)$ are those morphisms in $\CS(\Gr_R(X))$ which take the form $\varrho^R_m(X)^*\rho$ for $\rho \in \Hom(\varrho^R_{n\leq m}(X)^*\cs{L},\cs{L}')$ when $n\leq m$, and $\varrho^R_n(X)^*\rho$ for $\rho \in \Hom(\cs{L},\varrho^R_{m\leq n}(X)^*\cs{L}')$ when $m\leq n$.
\end{enumerate}
\end{remark}

@@

 \begin{remark}
In Section~\ref{ssec:CS_tori} we see that the \'etale site on $\Gr_R(X)$ is rich enough to geometrize all quasicharacters of $X(K)$ as $\ell$-adic local systems on $\Gr_R(X)$, where $X$ is a Neron model for an algebraic torus or an abelian variety over a local field $K$. 
It is natural to ask if the \'etale site on the generic fibre $X_K$ would have sufficed. 
This seems unlikely, since the geometric \'etale fundamental group of $\Gm{K}$ is ${\hat \ZZ}$, though limited results in this direction were established in \cite{Cunningham-Kamgarpour} when $K = \mathbb{Q}_p$.
\end{remark}

@@

where we refer to continuous cohomology, since $\Hom(T(K), \EEx)$ refers to continuous homomorphisms $T(K)\to \EEx$.
It now follows from the inflation-restriction exact sequence that the following diagram commutes.

@@

@@

Let $K'/K$ be a finite Galois extension of local fields and
let $k'/k$ be the corresponding extension of residue fields.
Let $R$ and $R'$ be the rings of integers of $K$ and $K'$, respectively.
Let $X_K \in \mathcal{N}$ have N\'eron model $X$, set $X_{K'} \ceq X_K \times_\Spec{K} \Spec{K'}$
and let $X'$ be a N\'eron model for $X_{K'}$.

>>>>

\begin{example}\label{eg:H2}
Consider the simplest non-trivial case, where $\bG = \{1, i, j, k\} \cong \ZZ/2\ZZ \times \ZZ/2\ZZ.$  Using \eqref{eq:H2comp},
we have $\Hh^2(\bG, \EEx) \cong \ZZ/2\ZZ$, on which $\Weil{}$ must act trivially, regardless of its action on $\bG$ itself.
The non-trivial element corresponds to the extension
\begin{equation} \label{eq:Qdef}
1 \to \EEx \to Q \to \bG \to 1,
\end{equation}
where $Q = \{c + c_ii + c_jj + c_kk \tq c,c_i,c_j,c_k \in \EE \mbox{ with exactly one nonzero}\}$ is a subgroup of the quaternion algebra over $\EE$.
Let $a$ be a $2$-cocycle corresponding to this extension, with values in $\{\pm 1\}$. When $\Frob{G}$ acts trivially on $\bG$,
any homomorphism $b : \bG \to \EEx$ will satisfy \eqref{eq:nohom}, and the corresponding $\alpha \oplus \beta$ are non-cohomologous in $\Hh^2(E_G^\bullet)$.
When $\Frob{G}$ exchanges $i$ and $j$, then we may take $b(1) = 1, b(i) = -1$ and $b(j) = b(k) = \pm 1$, up to coboundaries.
Finally, when $\Frob{G}$ cycles $i$, $j$ and $k$, any homomorphism $b : \bG \to \EEx$ will satisfy \eqref{eq:nohom},
but now the corresponding $\alpha \oplus \beta$ are all cohomologous in $\Hh^2(E_G^\bullet)$.  In each case, we may produce
an explicit character sheaf from the listed $a$ and $b$.

@@

Let $R$ be the ring of integers of $K$.
Write $\m$ for the maximal ideal of $R$ and set $R_n = R/\m^{n+1}$ for every non-negative integer $n$.
Suppose $X \in \mathcal{N}$.
Note that $X(K) = X(R)$.
A \emph{quasicharacter of $X(K)$} is a group homomorphism $X(K) \to \EEx$ that factors through $X(R) \to X(R_n)$ for some non-negative integer $n$.
We note that this definition is compatible with \cite{cassels-frohlich:AlgebraicNumberTheory}*{Ch XV, \S 2.3}.
The group of quasicharacters of $X(K)$ will be denoted by 
$\Hom_\text{}(X(K),\EEx)$
 and the subgroup of those that factor through $X(R_n)$ will be denoted by $\Hom_n(X(K),\EEx)$.
In this section we will see how to geometrize and categorify quasicharacters of $X(K)$ using character sheaves.

@@

For any $n\leq m$,  the surjective ring homomorphism $R_{m} \to R_n$ determines a
natural transformation 
\[
\varrho^R_{n\leq m} : \Gr^R_{m} \to \Gr^R_n
\]
between additive functors.
Crucially, $\varrho^R_{n\leq m}(X): \Gr^R_{m}(X)\to \Gr^R_n(X)$ is an affine morphism of $\Fq$-schemes, for every $R$-scheme $X$ and every $n\leq m$ \cite{bertrapelle-gonzales:Greenberg}*{Prop 4.3}.
This observation is key to the proof that, for any scheme $X$ over $R$, the projective limit 
\[
\Gr_R(X) \ceq \varprojlim_{n\in \NN} \Gr^R_n(X),
\]
taken with respect to the surjective morphisms $\varrho^R_{n\leq m}(X) : \Gr^R_{m}(X) \to \Gr^R_n(X)$,
exists in the category of group schemes over $\Fq$;
see \cite{EGAIV3}*{\S 8.2}.
This leads to the definition of the {\it Greenberg transform}:
\[
\Gr_R : \Sch{R} \to \Sch{\Fq}.
\]
By construction, the $\Fq$-scheme $\Gr_R(X)$ comes equipped with morphisms 
\[
\varrho^R_n(X) : \Gr^R(X) \to \Gr^R_n(X),\qquad \forall n\in \NN.
\]

@@

%Let $X \in \mathcal{N}$ be as above.
Set $S = \Spec{R}$ and $S_n = \Spec{R_n}$;
note that $S_0 = \Spec{\Fq}$ is the special fibre of $S$.
Let $X$ be a smooth commutative group scheme over $S$.
%We now define a category which will geometrize quasicharacters of $X(K)$.
For every non-negative integer $n$, the Greenberg transform $\Gr^R_n(X)$ is a {\it smooth} commutative group scheme over $S_0$.
%with component group scheme $\pi_0(X)_{\Fq} \ceq \pi_0(X) \times_S S_0$ and $\Fq$-rational points $X(R_n).$
%
The Greenberg transform $\Gr_R(X)$ of $X$ is a commutative group scheme over $\Fq$
%with component group scheme $\pi_0(X)_{\Fq}$
with $\Fq$-rational points $X(R)$.
%
The morphism of $\Fq$-schemes $\varrho^R_n(X) : \Gr_R(X) \to \Gr^R_n(X)$ induces a functor
\[
\varrho^R_n(X)^* : \CS(\Gr^R_n(X)) \to \CS(\Gr_R(X)),
\]
as in Lemma~\ref{lem:pullback}.
%This functor is faithful, so we may view $\CS(\Gr^R_n(X))$ as a subcategory of $\CS(\Gr_R(X))$.

@@

\begin{remark}
We offer the following alternate construction of $\QCS(X)$.
As above, let $X$ be a smooth group scheme over $R$.
Although $\Gr_R(X)$ is not locally of finite type and therefore not smooth,
let us consider the rigid monoidal category $\CS(\Gr_R(X))$ as defined in Section~\ref{ssec:category}, though without insisting that the commutative group $\Fq$-scheme $G$ is smooth.
%Let $X$ be a smooth commutative group scheme over $R$, with $R$ as above.
A quasicharacter sheaf for $X$ is an object of
the following rigid monoidal subcategory of $\CS(\Gr_R(X))$, denoted by $\QCS(X)$:
\begin{enumerate}
\item
objects in $\QCS(X)$ are the $\ell$-adic sheaves $\varrho^R_n(X)^*\cs{L}$, for $n\in \NN$ and $\cs{L} \in \CS(\Gr^R_n(X))$; 
\item
morphisms $\varrho^R_n(X)^*\cs{L} \to \varrho^R_m(X)^*\cs{L}'$ in $\QCS(X)$ are those morphisms in $\CS(\Gr_R(X))$ which take the form $\varrho^R_m(X)^*\rho$ for $\rho \in \Hom(\varrho^R_{n\leq m}(X)^*\cs{L},\cs{L}')$ when $n\leq m$, and $\varrho^R_n(X)^*\rho$ for $\rho \in \Hom(\cs{L},\varrho^R_{m\leq n}(X)^*\cs{L}')$ when $m\leq n$.
\end{enumerate}
\end{remark}

@@

 \begin{remark}
In Section~\ref{ssec:CS_tori} we see that the \'etale site on $\Gr_R(X)$ is rich enough to geometrize all quasicharacters of $X(K)$ as $\ell$-adic local systems on $\Gr_R(X)$, where $X$ is a Neron model for an algebraic torus or an abelian variety over a local field $K$. 
It is natural to ask if the \'etale site on the generic fibre $X_K$ would have sufficed. 
This seems unlikely, since the geometric \'etale fundamental group of $\Gm{K}$ is ${\hat \ZZ}$; however, limited results in this direction were established in \cite{Cunningham-Kamgarpour} when $K = \mathbb{Q}_p$.
\end{remark}

@@

where we refer to continuous cohomology, since $\Hom(T(K), \EEx)$ refers to continuous group homomorphisms $T(K)\to \EEx$.
It now follows from the inflation-restriction exact sequence that the following diagram commutes:

@@

 & 0.

@@

Let $K'/K$ be a finite Galois extension of local fields and
let $k'/k$ be the corresponding extension of residue fields.
Let $R$ and $R'$ be the rings of integers of $K$ and $K'$, respectively.
Suppose $X \in \mathcal{N}$ and set $X_{K'} \ceq X_K \times_\Spec{K} \Spec{K'}$
and let $X'$ be a N\'eron model for $X_{K'}$.
